Tool-driven Revolutions in Archaeological Science

### 2019-08-09 analysis/paper/jcaa-paper.Rmd
resolve merge conflict

# Revolutions in Science: Ideas-driven or Tools-driven?

Philosophers of science disagree whether revolutions in science are ideas-driven or tool-driven. One of the most widely-known models of ideas-driven change in science is Kuhn's [-@Kuhn_1962] effort to describe the history of science by proposing paradigm shifts resulting from revolutionary change in communities of researchers. In brief, this model proposes that the history of science consists of long periods of tradition-bound 'normal science' punctuated by short episodes of 'revolution'. Normal science was described by Kuhn as stagnant, routine day-to-day research focused on 'puzzle solving'. What counts as suitable puzzles, and acceptable solutions to these puzzles, is governed by the norms and procedures of the prevailing paradigm. Revolutions in science occur as punctuations in the equilibria of normal science, when anomalous observations culminate in changes replacing one paradigm with another. In this conversion or speciation process the community adopts a new way of doing science. There has been much debate about how to draw a distinction between so-called normal and revolutionary science. [@watkins1970against; @toulmin1970does; @Casadevalle00158_16]. A notable criticism is that Kuhn's normal science is unscientific, as it describes a situation where critical science had contracted into defensive metaphysics resulting from the domination of a ruling dogma [@popper1970]. This is at odds with a view of science as a continuous state of evolution, with researchers simultaneously employing many styles of thinking and doing science, and continuously confronting theory with evidence and modifying their ideas based on the outcome. In this view, based on the post-World War II era of science, radical discontinuities are rare, and major developments instead emerge from the division and recombination of already mature fields [@andersen2013second]. Given the rarity of radical discontinuities, some biologists have argued that change in biology is better characterised by new ideas (e.g. Mendelian heredity, Darwinian evolution, and molecular genetics) replacing not a former paradigm, but a conceptual vacuum [@wilkins1996are].

In archaeology we have seen claims for paradigm shifts in several areas. The most wide-ranging application is Clark's [-@Clark_1993] comparison of North American archaeology to European archaeology during the 1970s-1990s. His analysis describes a multi-paradigm situation where many research communities (i.e. North American and European) simultaneously operate in different paradigms (which Kuhn noted as a sign of a developing or immature discipline). Clark's account is not a complete Kuhnian analysis because it lacks a revolutionary change event where one paradigm is replaced by another, although he hints that processual archaeology may be considered a radical discontinuity relative to the culture-historical approaches that preceded it [@Clark_1993, pp. 206].

@@

We also see mentions of paradigm shifts in archaeology in reference to the shift from processual to post-processual archaeology [@Härke_2002; @Koerner_2018], as well as more localised and specific shifts in archaeological practice and thinking. For example, @Snodgrass_2002 claims that among Classical Archaeologists, an increase in research on previously neglected periods of antiquity, such as the Greek Early Iron Age, constitutes a paradigm shift. @McAnany_Rowe_2015 propose that the appearance of community-based participatory models of research among some communities of archaeologists is a paradigm shift, though they concede that it is transformational rather than revolutionary. @harris2012interfacing similarly argues that community crowdsourced geographic knowledge (or volunteered geographic information) could be paradigm shift for archaeological communities. Archaeologists using LIDAR (light detection and ranging) for remote geospatial imaging of cultural landscapes have claimed the application of this technology to archaeology is a paradigm shift [@chase2012geospatial; @howey2016detecting]. @Fuller_2010 has used the term to describe a shift in thinking from the emergence of agriculture as a  'Neolithic Revolution' to a protracted and entangled process, happening several times independently.
Most of these claims for  Kuhnian paradigm shifts differ in meaning, which highlights an important complication with using Kuhn's concept of a paradigm-shift, namely the multiple definitions of a paradigm in his writings.

@@

We focus here on the third factor, free and open source software, because it is the most generic factor and so the most relevant beyond ecology. Archaeology and ecology each have a high diversity of data collection methods, and types of data analyses they conduct. An important similarlity for the two disciplines is working with the field-collected data on a computer to prepare it for publication. R is a widely used free and open source data analysis tool in many research communities [@Tippmann_2015; @Baker_2017; @Thieme_2018]. It is the dominant programming language in archaeology and ecology, so it is a good proxy for the adoption of open source scientific programming languages in these disciplines. To investigate changes in the use of R in archaeology and ecology, we obtained reference lists from a sample of scholarly articles in the Web of Science database and examined patterns in the citation of the R program over time.

>>>>

# Revolutions in Science: Ideas-driven or Tool-driven?

Philosophers of science disagree whether revolutions in science are ideas-driven or tool-driven. One of the most widely-known models of ideas-driven change in science is Kuhn's [-@Kuhn_1962] effort to describe the history of science by proposing paradigm shifts resulting from revolutionary change in communities of researchers. In brief, this model proposes that the history of science consists of long periods of tradition-bound 'normal science' punctuated by short episodes of 'revolution'. Normal science was described by Kuhn as stagnant, routine day-to-day research focused on 'puzzle solving'. What counts as suitable puzzles, and acceptable solutions to these puzzles, is governed by the norms and procedures of the prevailing paradigm. Revolutions in science occur as punctuations in the equilibria of normal science, when anomalous observations culminate in changes replacing one paradigm with another in a conversion or speciation process where the community adopts a new way of doing science. There has been much debate about how to draw a distinction between so-called normal and revolutionary science. [@watkins1970against; @toulmin1970does; @Casadevalle00158_16]. A notable criticism is that Kuhn's normal science is unscientific, a situation where critical science had contracted into defensive metaphysics resulting from the domination of a ruling dogma [@popper1970]. This is at odds with a view of science as a continuous state of evolution, with researchers simultaneously employing many styles of thinking and doing science, and continuously confronting theory with evidence and modifying their ideas based on the outcome. In this view, based on the post-World War II era of science, radical discontinuities are rare, and major developments instead emerge from the division and recombination of already mature fields [@andersen2013second]. Given the rarity of radical discontinuities, some biologists have argued that change in biology is better characterised by new ideas (e.g. Mendelian heredity, Darwinian evolution, and molecular genetics) replacing not a former paradigm, but a conceptual vacuum [@wilkins1996are].

In archaeology we have seen claims for paradigm shifts in several areas. The most wide-ranging application is Clark's [-@Clark_1993] comparison of North American archaeology to European archaeology during the 1970s-1990s. Clark, similarly to Binford and Sabloff [-@binford1982paradigms], identified the North American archaeological paradigm as rooted in anthropology, founded on post-positivist biases, and distinctive with its critical-realist ontology, a modified-objectivist epistemology, and an experimental-manipulative method. In contrast, the European archaeological paradigm is described a kind of history, mostly in a strict empiricist tradition, and typified by a realist ontology, an inductivist epistemology, and by an observational method. Clark's analysis describes a multi-paradigm situation where many research communities (i.e. North American and European) simultaneously operate in different paradigms (which Kuhn noted as a sign of a developing or immature discipline). Clark's account is not a complete Kuhnian analysis because it lacks a revolutionary change event where one paradigm is replaced by another, although he hints that processual archaeology may be considered a radical discontinuity relative to the culture-historical approaches that preceded it [@Clark_1993, pp. 206].

@@

We also see mentions of paradigm shifts in archaeology in reference to the shift from processual to post-processual archaeology [@Härke_2002; @Koerner_2018], as well as more localised and specific shifts in archaeological practice and thinking. For example, @Snodgrass_2002 claims that among Classical Archaeologists, an increase in research on previously neglected periods of antiquity, such as the Greek Early Iron Age, constitutes a paradigm shift. @McAnany_Rowe_2015 propose that the appearance of community-based participatory models of research among some communities of archaeologists is a paradigm shift, though they concede that it is transformational rather than revolutionary. @harris2012interfacing similarly argues that community crowdsourced geographic knowledge (or volunteered geographic information) could be paradigm shift for archaeological communities. Archaeologists using LIDAR (light detection and ranging) for remote geospatial imaging of cultural landscapes have claimed the application of this technology to archaeology is a paradigm shift [@chase2012geospatial; @howey2016detecting]. @Fuller_2010 has used the term to describe a shift in thinking from the emergence of agriculture as a parallel, simultaneous event at the start of the Holocene (the Neolithic Revolution), to its emergence as a protracted and entangled process, happening 20 or so times independently, following different pathways. Most of these claims for paradigm shifts differ in meaning from Kuhn's classical formulation, and this highlights an important complication with using Kuhn's concept of a paradigm-shift, namely the multiple definitions of a paradigm in Kuhn's writings.

@@

We focus here on the third factor, free and open source software, because it is the most generic factor and so the most relevant beyond ecology. Archaeology and ecology each have a high diversity of data collection methods, and types of data analyses they conduct. An important similarlity for the two disciplines is working with the field-collected data on a computer to prepare it for publication. R is a widely used free and open source data analysis tool in many research communities [@Tippmann_2015; @Baker_2017; @Thieme_2018]. Our observations suggest it is the dominant programming language in archaeology and ecology, so it is a good proxy for the adoption of open source scientific programming languages in these disciplines. To investigate changes in the use of R in archaeology and ecology, we obtained reference lists from a sample of scholarly articles in the Web of Science database and examined patterns in the citation of the R program over time.
